---
title: "Tornadoes Cause Most Damage to People and Property in the United States"
output: 
  html_document:
    keep_md: true
---

#Synopsis
The weather event that is most harmful in relation to population health in the US is Tornadoes. It causes by far the most injuries and fatalities. The next most detrimental event is Excessive Heat, which produces slightly more fatalities than the cluster of the remaining 5 Events (Flash Flood, Flood, Heat, Lightning, Thunderstorm Wind).

The weather event that is most economically damaging is also Tornados, beating out the next closest event by about $ 1.5 million. The next two most damaging events are Flash Floods and TSTM Wind. If all Thunderstorm Wind categories are combined, they would contribute to the 2nd most Property Damaging Event.

Overall, the most damaging event in the US in relation to people and property damage is Tornadoes. 


#Data Processing
###Requirements to load data:
- R version 3.4.1
- data file named "repdata%2Fdata%2FStormData.csv.bz2" in your working directory

```{r cache=TRUE}
library(ggplot2)

data <- read.csv("repdata%2Fdata%2FStormData.csv.bz2", header=TRUE, sep=",")
```

#Results Part 1 - Which EVTYPE (Event Type) is Most Harmful to Health?
This was interpreted as which Event Type creates the most fatalities and injuries.  

First, sum up total Fatalites by Event Type, and total Injuries by Event type.

```{r}
sumPeople <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data, sum)

```

Next, combine both data sets, sort FATALITIES for Top 5

```{r}
fatINJor <- sumPeople[order(sumPeople$FATALITIES, sumPeople$INJURIES),]
```

Top 5 Events for FATALITIES
```{r}
tail(fatINJor, n=5)
```

Top 5 Events for INJURIES
```{r}
fatINJor <- sumPeople[order(sumPeople$INJURIES, sumPeople$FATALITIES),]
tail(fatINJor, n=5)
```

Subset original data to include Top 7 Event Types for BOTH Fatalites and Injuries, then find Total Fatalities and Injuries by Event Type
```{r}
topsev <- subset(data, data$EVTYPE == "TORNADO" | data$EVTYPE == "TSTM WIND" | data$EVTYPE == "FLOOD" | data$EVTYPE == "EXCESSIVE HEAT" | data$EVTYPE == "LIGHTNING" | data$EVTYPE == "FLASH FLOOD" | data$EVTYPE == "HEAT")
```

Find Totals for FATALITIES & INJURIES
```{r}
topsevTotal <- aggregate(cbind(FATALITIES, INJURIES)~EVTYPE, topsev, sum)
topsevTotal
```

Plot Fatalities VS Injuries by Event Type
```{r}
library(ggplot2)
qplot(FATALITIES, INJURIES, data=topsevTotal, color=EVTYPE, main="Fatalities And Injuries by 7 Weather Events in US")
ggsave("people.png")
dev.off()
```



#Results Part 2 - Which EVTYPE (Event Type) Has Most Economic Consequences?
Sum up the Property Damage amounts by Event Type, and reorder
```{r}
damsum <- aggregate(PROPDMG ~ EVTYPE, data, sum)

damsum <- damsum[order(damsum$PROPDMG), ]
tail(damsum, n=10)
```
Subset top 10 Economically Damaging Events (> PROPDMG >= 13,000)

```{r}
damSumg13 <- subset(damsum, PROPDMG >=130000)
```

Plot Total Property Damage by Top 10 Weather Events

```{r}
g <- ggplot(damSumg13, aes(EVTYPE, PROPDMG))
g+geom_point(aes(color = EVTYPE))+
  theme(axis.text.x=element_blank())+      
  labs(x="Weather Event", y="Monetary Property Damage") + 
  ggtitle("Monetary Property Damage for US by Top 10 Weather Events")+
  ylim(c(0,3300000))
ggsave("money.png")
dev.off()
 ```

